Penalized linear mixed models for structured genetic data

### 2021-03-19 submissions/response/response.tex

  We have added a new section (Section 3.8) that specifically addresses the availability of software implementing these methods.  As far as whether the software is ``simple'', we tried to make our R package, \texttt{penalizedLMM}, as user-friendly as possible, with convenient functions for importing data from PLINK files and preprocessing (e.g., carrying out genomic imputation prior to fitting).  We hope that this makes these models relatively easy to fit and use for practitioners.

\item \emph{I understand that the authors’ focus is on linear model. But, many if not most genetic studies do not have a continuous outcome and linear model. Can the models/approaches be extended to other outcomes/models with reasonable effort? Either way, it is worth knowing.}

  This is an excellent point.  We have expanded our discussion of this issue in section 2.3, which now contains the following paragraph:

  \begin{quote}
    In many genetic association studies the phenotype of interest is binary, for example the presence or absence of a heritable trait. Unfortunately, the extension of LMMs and penalized LMMs to nonlinear models is not straightforward. As we explain in Section 3, LMMs and penalized LMMs are able to efficiently correct for dependencies among subjects by preprocessing the data by means of a weighted rotation. This procedure does not directly generalize beyond the linear regression case. Other approaches are possible, for example Mandt et al. (2017) generalize the penalized LMM modeling paradigm to binary phenotypes by thresholding the outcome variable through a probit likelihood. However, there is no work that we are aware of that explicitly addresses the use of such models for GWAS. We hope that this situation changes in the near future.
  \end{quote}

### 2021-03-19 main.tex
updated 2-panel figure 5. updating figure refs, additional figure numbers

\begin{equation}
\centering
\includegraphics{figures/dag1.eps}
\label{diag:ps_env}
\end{equation}

@@

\begin{equation}
\centering
\includegraphics{figures/dag2.eps}
\label{diag:ps_env_block}
\end{equation}

@@

\begin{equation}
\centering
\includegraphics{figures/dag3.eps}
\label{diag:ps_env_block2}
\end{equation}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/figure_01.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with 4 subpopulations. Mean and standard error (SE) bars displayed for 1000 replications.}
    \label{fig:mse}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/figure_02.png}
\caption{Absolute allele frequency differences of selected, non-causal SNPs across methods, in the presence and absence of environmental confounding for two Empirical subpopulations, $\eta = 0.5$, 1000 replications.}
\label{fig:pop_inf}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/figure_03.png}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations, $\eta = 0.5$, 1000 replications.}
    \label{fig:big_vs_small}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/figure_04.png}
     \caption{Estimation accuracy and computing time of LMM-lasso-Rakitsch and LMM-lasso-ggmix for $\bbeta$ and $\eta$ for varying $\eta$ levels and sample sizes with Independent Subpopulations data, coarse subpopulation structure, $\xi = 0.8$, 1000 replications.}
    \label{fig:eta_beta_mse}
\end{figure}

@@

Selection accuracy was evaluated using precision. All $\beta > 0$ and we calculate precision = $\#\{ \beta > 0 \cap \widehat{\beta} > 0  \}/\#\{ \widehat{\beta} \ne 0 \}$ in order to ensure SNPs were not only selected, but their directions estimated correctly. \anna{Figure~\ref{fig:gw_prc}} displays each method's precision for the corresponding number of selected SNPs. PC-lasso and LMM-lasso consistently outperform the ordinary lasso in selecting causal SNPs, particularly for the first $\approx$ 100 SNPs each method allows into the model. For the most part, PC-lasso outperforms LMM-lasso in this setting as well. None of the method are able to accurately select more than about 100 SNPs.  However, because the ordinary lasso does not account for environmental confounding, it cannot reliable select \emph{any} SNPs: the first 38 SNPs selected were all false positives.

Estimation accuracy was assessed via MSE, as defined in Section~\ref{sec:sim-mse}. \anna{Figure~\ref{fig:gw_mse}} displays this value for each method and number of selected SNPs. In accordance with previous simulation results in Section~\ref{sec:sim-mse}, the ordinary lasso results in biased estimates compared to PC and LMM-lasso. This bias is most dramatic for the first $\approx$ 100 SNPs selected by each method. For any number of selected SNPs, however, both PC-lasso and LMM-lasso produce much more accurate estimates than the ordinary lasso.

\anna{Make sure this gets an overall figure number}
\begin{figure}[H]
     \centering
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05a.pdf}
         \caption{Precision for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_prc}
     \end{subfigure}
    \hfill
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05b.pdf}
         \caption{MSE for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_mse}
     \end{subfigure}
     \begin{subfigure}[b]{0.15\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05c.pdf}
     \end{subfigure}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/figure_07.png}
\caption{Absolute correlation of selected, non-causal SNPs with the first principal component (PC1) of the Empirical genome-wide data for each method and number of selected SNPs. The plotted line is the mean absolute correlation of all incorrectly selected SNPs for a particular selection number, with shaded region indicating SE bounds. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_pc1_corr}
\end{figure}

>>>>

\begin{equation}
\centering
\includegraphics{figures/dag1.pdf}
\label{diag:ps_env}
\end{equation}

@@

\begin{equation}
\centering
\includegraphics{figures/dag2.pdf}
\label{diag:ps_env_block}
\end{equation}

@@

\begin{equation}
\centering
\includegraphics{figures/dag3.pdf}
\label{diag:ps_env_block2}
\end{equation}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/figure_01.pdf}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with 4 subpopulations. Mean and standard error (SE) bars displayed for 1000 replications.}
    \label{fig:mse}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/figure_02.pdf}
\caption{Absolute allele frequency differences of selected, non-causal SNPs across methods, in the presence and absence of environmental confounding for two Empirical subpopulations, $\eta = 0.5$, 1000 replications.}
\label{fig:pop_inf}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/figure_03.pdf}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations, $\eta = 0.5$, 1000 replications.}
    \label{fig:big_vs_small}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/figure_04.pdf}
     \caption{Estimation accuracy and computing time of LMM-lasso-Rakitsch and LMM-lasso-ggmix for $\bbeta$ and $\eta$ for varying $\eta$ levels and sample sizes with Independent Subpopulations data, coarse subpopulation structure, $\xi = 0.8$, 1000 replications.}
    \label{fig:eta_beta_mse}
\end{figure}

@@

Selection accuracy was evaluated using precision. All $\beta > 0$ and we calculate precision = $\#\{ \beta > 0 \cap \widehat{\beta} > 0  \}/\#\{ \widehat{\beta} \ne 0 \}$ in order to ensure SNPs were not only selected, but their directions estimated correctly. Figure~\ref{fig:gw_prc_mse}(a) displays each method's precision for the corresponding number of selected SNPs. PC-lasso and LMM-lasso consistently outperform the ordinary lasso in selecting causal SNPs, particularly for the first $\approx$ 100 SNPs each method allows into the model. For the most part, PC-lasso outperforms LMM-lasso in this setting as well. None of the method are able to accurately select more than about 100 SNPs.  However, because the ordinary lasso does not account for environmental confounding, it cannot reliably select \emph{any}: the first 38 SNPs selected were all false positives.

In addition to selection, estimation accuracy was assessed via MSE, as defined in Section~\ref{sec:sim-mse}. Figure~\ref{fig:gw_prc_mse}(b) displays this value for each method and number of selected SNPs. In accordance with previous simulation results in Section~\ref{sec:sim-mse}, the ordinary lasso results in biased estimates compared to PC and LMM-lasso. This bias is most dramatic for the first $\approx$ 100 SNPs selected by each method. For any number of selected SNPs, however, both PC-lasso and LMM-lasso produce much more accurate estimates than the ordinary lasso.

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/figure_05.pdf}
\caption{Precision and MSE for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_prc_mse}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/figure_06.pdf}
\caption{Absolute correlation of selected, non-causal SNPs with the first principal component (PC1) of the Empirical genome-wide data for each method and number of selected SNPs. The plotted line is the mean absolute correlation of all incorrectly selected SNPs for a particular selection number, with shaded region indicating SE bounds. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_pc1_corr}
\end{figure}
